A primer to numerical simulations: The perihelion motion of Mercury

### 2018-01-09 tex/MercuryNumerical.tex
updated figures

In \python{}, this reads
\begin{lstlisting}
# Definition of parameters
rM0 = 4.6    # in units of R0
vM0 = 0.51   # in units of R0/T0
c_a = 1.01   # in units of R0/T0**2
TM  = 88.    # in units of T0
rS  = 3.0e-7 # in units of R0
a2  = 8.2e-7 # in units of R0**2 \ ,
\end{lstlisting}
where the last two quantities refer to the Schwarzschildradius of the sun and the parameter $a^2$ defined in Eqs. (\ref{rsdef}) and (\ref{a2def}), respectively.
So far we only fixed the length of the vectors.
Next, we set up the initial directions, which will describe the motion in the two-dimensional plane.  We will build on the existing \python{} module \texttt{VPython}, which provides an implementation for treating vectors as well as their visualization.
The first object of interest is a \texttt{vector}, which takes three-dimensional coordinates as its input.
Because of our choice of initial conditions (we picked the initial vectors in the perihelion), the velocity of Mercury is perpendicular to the vector which connects Mercury and sun (see figure \ref{fig:sun_merc}):
\begin{lstlisting}
# Import the class vector from vpythons module visual
from vpython import vector
# Initialize distance and velocity vectors
vec_rM0 = vector(0, rM0, 0)
vec_vM0 = vector(vM0, 0, 0) \ .
\end{lstlisting}
To fix the time step $\Delta t$ we use the estimate of Eq.~(\ref{eq:check}) as a guidance to write
\begin{lstlisting}
# Definition of the time step
dt = 2 * vM0 / c_a / 20 \ .
\end{lstlisting}
Here the factor $1/20$ makes sure that $\Delta t$ is indeed consistent with Eq.~(\ref{eq:check}).
Now we are in the position to calculate location and velocity of the planet at $t_0+\Delta t$ using the following equations\footnote{%
	Note, that there are different notions for numerically integrating differential equations with different accuracies.
	Thus, the ordering and exact equations for updating $a_{MS}$, $v_M$ and $r_{MS}$ are not unique, but all correct prescriptions lead to the same result for sufficiently small $\Delta t$.
	For this particular problem, the following definition is in good balance between numerically stable and inexpensive to compute.
}%
\begin{lstlisting}
# Compute the strength of the acceleration
temp = 1 + alpha * rS / vec_rM_old.mag + beta * a2 / vec_rM_old.mag**2
aMS  = c_a * temp / vec_rM_old.mag**2
# Multiply by the direction
vec_aMS = - aMS * ( vec_rM_old / vec_rM_old.mag )
# Update velocity vector
vec_vM_new = vec_vM_old + vec_aMS * dt
# Update position vector
vec_rM_new = vec_rM_old + vec_vM_new * dt
\end{lstlisting}
Note basic vector operations are already implemented in the predefined \texttt{vector} class.
The difference and sum of two vectors, or the scalar vector multiplication return vectors themselves.
Also the magnitude of a vector --- \texttt{vector.mag} --- is a property of the vector and can be easily extracted.

@@

As an example we  extract the perihelion motion using the parameters provided by General Relativity, namely $\alpha = 0$ and $\beta = 3$.
Considering one result from the numeric simulation, e.g., $\delta\Theta(\alpha = 0, \beta=10^6) \approx 10.7^{\degree}$, the sought-after angle can be calculated by a linear interpolation as
\begin{equation}
	\delta\Theta (\alpha=0, \beta=3) = \frac{10.6^{\degree}}{10^6} \times 3 = 0.116''
	\, .
\end{equation}
Depending on the knowledge of the students, multiple points and also estimated numerical uncertainties could be used to extract this value by a linear regression.
Note that, when comparing the results of the simulation with the experimental result, one Earth year corresponds to $T_E/T_M \approx 365d/88d\approx 4.15$ Mercury years.
Therefore, the perihelion motion per 100 earth years is $0.116''\cdot 415\approx 48.2''$.
This is close to experimental value of $\delta\Theta_{\mathrm{exp}} = 43.0''$.

>>>>

In \python{}, this reads
\begin{lstlisting}
# Definition of parameters
rM0 = 4.6    # in units of R0
vM0 = 0.51   # in units of R0/T0
c_a = 1.01   # in units of R0/T0**2
TM  = 88.    # in units of T0
rS  = 3.0e-7 # in units of R0
a2  = 8.2e-7 # in units of R0**2
\end{lstlisting}
where the last two quantities refer to the Schwarzschildradius of the sun and the parameter $a^2$ defined in Eqs. (\ref{rsdef}) and (\ref{a2def}), respectively.
So far we only fixed the length of the vectors.
Next, we set up the initial directions, which will describe the motion in the two-dimensional plane.  We will build on the existing \python{} module \texttt{VPython}, which provides an implementation for treating vectors as well as their visualization.
The first object of interest is a \texttt{vector}, which takes three-dimensional coordinates as its input.
Because of our choice of initial conditions (we picked the initial vectors in the perihelion), the velocity of Mercury is perpendicular to the vector which connects Mercury and sun (see figure \ref{fig:sun_merc}):
\begin{lstlisting}
# Import the class vector from vpythons module visual
from vpython import vector
# Initialize distance and velocity vectors
vec_rM0 = vector(0, rM0, 0)
vec_vM0 = vector(vM0, 0, 0)
\end{lstlisting}
To fix the time step $\Delta t$ we use the estimate of Eq.~(\ref{eq:check}) as a guidance to write
\begin{lstlisting}
# Definition of the time step
dt = 2 * vM0 / c_a / 20
\end{lstlisting}
Here the factor $1/20$ makes sure that $\Delta t$ is indeed consistent with Eq.~(\ref{eq:check}).
Now we are in the position to calculate location and velocity of the planet at $t_0+\Delta t$ using the following equations\footnote{%
	Note, that there are different notions for numerically integrating differential equations with different accuracies.
	Thus, the ordering and exact equations for updating $a_{MS}$, $v_M$ and $r_{MS}$ are not unique, but all correct prescriptions lead to the same result for sufficiently small $\Delta t$.
	For this particular problem, the following definition is in good balance between numerically stable and inexpensive to compute.
}%
\begin{lstlisting}
# Compute the strength of the acceleration
temp = 1 + alpha * rS / vec_rM_old.mag + beta * a2 / vec_rM_old.mag**2
aMS  = c_a * temp / vec_rM_old.mag**2
# Multiply by the direction
vec_aMS = - aMS * ( vec_rM_old / vec_rM_old.mag )
# Update velocity vector
vec_vM_new = vec_vM_old + vec_aMS * dt
# Update position vector
vec_rM_new = vec_rM_old + vec_vM_new * dt
\end{lstlisting}
Note basic vector operations are already implemented in the predefined \texttt{vector} class.
The difference and sum of two vectors, or the scalar vector multiplication return vectors themselves.
Also the magnitude of a vector --- \texttt{vector.mag} --- is a property of the vector and can be easily extracted.

@@

As an example we  extract the perihelion motion using the parameters provided by General Relativity, namely $\alpha = 0$ and $\beta = 3$.
Considering one result from the numeric simulation, e.g., $\delta\Theta(\alpha = 0, \beta=10^6) \approx 10.7^{\degree}$, the sought-after angle can be calculated by a linear interpolation as
\begin{equation}
	\delta\Theta (\alpha=0, \beta=3) = \frac{10.7^{\degree}}{10^6} \times 3 = 0.116''
	\, .
\end{equation}
Depending on the knowledge of the students, multiple points and also estimated numerical uncertainties could be used to extract this value by a linear regression.
Note that, when comparing the results of the simulation with the experimental result, one Earth year corresponds to $T_E/T_M \approx 365d/88d\approx 4.15$ Mercury years.
Therefore, the perihelion motion per 100 earth years is $0.116''\cdot 415\approx 48.2''$.
This is close to experimental value of $\delta\Theta_{\mathrm{exp}} = 43.0''$.
